Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-24 paper.Rmd
edits

Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. It does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g. @goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g. @beger:etal:2020.] B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation. These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.^[We also note there are additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model is not the best; it actually performs worse than the atheoretical, ktchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split. 
The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015;@beger:etal:2017]. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated  on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B&S Table 2.

@@

> Figure 1 displays the corresponding ROC curves, smoothed for ease of interpretation.

This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results.
Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves *prior* to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.  Thus, smoothing changes the ordering of alternatives. It is not a cosmetic change as suggested.

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC. Four of the eight show a small positive bias from smoothing, while the other four display a small negative bias. The magnitudes of these are not close to the level of positive bias found in the escalation model.

@@

The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguably performs better still.

@@

With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives). With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses. Fifteen to 16 civil war onsets in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war onsets. Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e. counting ongoing civil wars as if they were continuing onsets.

@@

\begin{table}
\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
\centering
\begin{tabular}{ll|c|c|l}
\multicolumn{5}{c}{\textbf{Original confusion matrices with civil war \textit{incidence}}} \\
% BS, assuming persistence
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 132 & 17 \\
\cline{3-4}
& $Y=1$ & 2 & 13 \\
\cline{3-4}
% BS, assuming change
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 132 & 16 \\
\cline{3-4}
& $Y=1$ & 2 & 14 \\
\cline{3-4}
\\ \midrule
\multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
% fixed, assuming persistence
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 134 & 30 \\
\cline{3-4}
& $Y=1$ & 0 & 0 \\
\cline{3-4}
% fixed, assuming change
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 134 & 28 \\
\cline{3-4}
& $Y=1$ & 0 & 2 \\
\cline{3-4}
\end{tabular}
\end{table}

The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}. In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1. As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives. This is the norm when predicting rare events, and thus not unusual at all. The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding. This is, at face value, a credit to the model's accuracy.

@@

However, for many, a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables. Typically it is fused together from a meta-analysis of prior work on the topic under consideration. We call these types of theory weak theory.
They are weak because they do not include two characteristics of strong theory, though they may include one:

@@

\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and
government is the target and typically involve appeals for political or institutional
reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which government
is the source and either opposition or rebels is the target. Nonviolent
repression includes threats, sanctions, restrictions on freedom of movement, or
rejections of group demands.
\item Low-level violence includes all events in which
(1) government is the source and opposition is the target, (2) opposition is the
source and government is the target, (3) government is the source and rebels are
the target, or (4) rebels are the source and government is the target.
\end{itemize}

@@

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible it is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].

>>>>

Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed, they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. It does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives. They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g., @goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g., @beger:etal:2020.] B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation. These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.^[We also note there are additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

<!-- [RM: I'm not sure what this sentence is doing in this paragraph. It was the second sentence in the paragraph below] The data in rare events problems, like this one, restrict the number of true positive rate values and lead to non-continuous ROC curves;  -->

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015;@beger:etal:2017]. Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees. Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new random forest model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split. The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B&S Table 2.

@@

> Figure 1 displays the corresponding ROC curves, _smoothed for ease of interpretation_. (emphasis added)

This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results. Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves *prior* to AUC calculation.

ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, for that matter, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way. Thus, smoothing changes the ordering of alternatives. It is not a cosmetic change as suggested.

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average. The Escalation model, however, _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values \textbf{only} benefits the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC. Four of the eight show a small positive bias from smoothing, while the other four display a small negative bias. The magnitudes of these are not close to the level of positive bias found in the escalation model.

@@

The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model performs better still.

@@

With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives). With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, they report that there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses. Fifteen to 16 civil war _onsets_ in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war _onsets_. Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e. counting ongoing civil wars as if they were onsets.

@@

\begin{table}
\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
\centering
\begin{tabular}{ll|c|c|l}
\multicolumn{5}{c}{\textbf{Original confusion matrices with civil war \textit{incidence}}} \\
% BS, assuming persistence
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 132 & 17 \\
\cline{3-4}
& $Y=1$ & 2 & 13 \\
\cline{3-4}
% BS, assuming change
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 132 & 16 \\
\cline{3-4}
& $Y=1$ & 2 & 14 \\
\cline{3-4}
\\ \midrule
\\
\multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
% fixed, assuming persistence
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 134 & 30 \\
\cline{3-4}
& $Y=1$ & 0 & 0 \\
\cline{3-4}
% fixed, assuming change
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 134 & 28 \\
\cline{3-4}
& $Y=1$ & 0 & 2 \\
\cline{3-4}
\end{tabular}
\end{table}

The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}. In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1, according to the civil conflict dataset they rely on. As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives. This is the norm when predicting rare events, and thus not unusual at all. The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding. This is, at face value, a credit to the model's accuracy.

@@

However, for many, a theory is simply a falsifiable story or narrative that makes causal claims about the relationships among variables. Typically it is fused together from a meta-analysis of prior work on the topic under consideration. We call these types of theory weak theory. They are weak because they do not include two characteristics of strong theory, though they may include one:

@@

\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and government is the target and typically involve appeals for political or institutional reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which the government is the source and either opposition or rebels is the target. Nonviolent repression includes threats, sanctions, restrictions on freedom of movement, or rejections of group demands.
\item Low-level violence includes all events in which (1) the government is the source and opposition is the target, (2) the opposition is the source and government is the target, (3) the government is the source and rebels are the target, or (4) rebels are the source and government is the target.
\end{itemize}

@@

Indeed, most empirical social science research follows this approach: (1) delineation of an argument, (2) selection of variables that represent a plausible or compelling representation of the concepts, followed by (3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story.

A strong theory gives more guidance on compelling choices than does a weak theory. Accordingly, while the choice is plausible it is not necessarily compelling. Another choice might be equally plausible. In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups. In practical terms there is a wide array of choices that are available to instantiate any particular theory. A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics. One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

At this level of detail, the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].
